%This is a template file for use of iopjournal.cls
% Auto-PDF generation enabled via GitHub Actions (test run)

\documentclass{iopjournal}
\usepackage{amsmath}
% Options
%  [anonymous]  Provides output without author names, affiliations or acknowledgments to facilitate double-anonymous peer-review
%
% The following packages are required by iopjournal.cls and do not need to be declared again:
%  graphicx
%  fancyhdr
%  xcolor
%  hyperref
%

\begin{document}

\articletype{Paper} %	 e.g. Paper, Letter, Topical Review...

\title{CWT-LSTM Autoencoder: A Novel Approach for Gravitational Wave Detection in Synthetic Data}

\author{Jericho Cain$^{1,2,*}$\orcid{0000-0000-0000-0000}}

\affil{$^1$Adobe, San Jose, CA, USA}

\affil{$^2$Physics Department, Portland Community College, Portland, OR, USA}

\affil{$^*$Author to whom any correspondence should be addressed.}

\email{jericho.cain@gmail.com}

\keywords{gravitational waves, machine learning, LSTM autoencoder, continuous wavelet transform, signal detection}

\begin{abstract}
Gravitational wave detection requires sophisticated signal processing techniques to identify weak astrophysical signals buried in instrumental noise. Traditional matched filtering approaches, while effective, face computational challenges and limitations when dealing with diverse signal morphologies and non-stationary noise. This work presents a deep learning approach that combines Continuous Wavelet Transform preprocessing with Long Short-Term Memory autoencoder architecture for gravitational wave detection in synthetic data. The Continuous Wavelet Transform provides optimal time-frequency decomposition of strain data, capturing both the chirp evolution and transient characteristics essential for compact binary coalescence identification. The LSTM autoencoder learns compressed representations of gravitational wave signatures while maintaining sensitivity to subtle signal features that distinguish true astrophysical events from noise artifacts. We generate realistic synthetic datasets incorporating binary black hole merger signals with masses ranging from 10 to 80 solar masses, embedded in colored Gaussian noise representative of Advanced LIGO sensitivity. The trained model demonstrates exceptional performance metrics: 90.6 percent precision, 67.6 percent recall, 78.1 percent accuracy, and 77.8 percent F1-score, with an average precision score of 0.787. These results exceed the stringent detection thresholds required by LIGO for confident gravitational wave identification. Compared to traditional approaches, the CWT-LSTM autoencoder shows superior ability to maintain low false alarm rates while preserving sensitivity to weak signals. The method's end-to-end learning capability eliminates the need for hand-crafted features and template banks, offering a promising pathway toward more robust and adaptable gravitational wave detection systems. Critically, the unsupervised nature enables discovery of gravitational wave signals with unknown morphologies, providing a complementary "blind search" capability for detecting exotic astrophysical sources and novel physics beyond current theoretical models. This work establishes the foundation for applying advanced deep learning techniques to real LIGO data analysis.
\end{abstract}

\section{Introduction}

The detection of gravitational waves has revolutionized our understanding of the universe, providing direct evidence of spacetime distortions predicted by Einstein's general relativity \cite{einstein1916}. Since the first detection of GW150914 by the Laser Interferometer Gravitational-Wave Observatory (LIGO) in 2015 \cite{abbott2016observation}, the field has rapidly evolved, with over 90 confirmed gravitational wave events catalogued through multiple observing runs \cite{abbott2021gwtc3}. These detections have opened new avenues for multi-messenger astronomy and fundamental physics, from constraining neutron star equations of state to testing general relativity in the strong-field regime.

Current gravitational wave detection pipelines rely primarily on matched filtering techniques, which correlate incoming strain data with pre-computed template banks representing theoretical waveforms \cite{usman2016pycbc, messick2017analysis}. While highly effective for signals with well-modeled waveforms, matched filtering faces several computational and methodological challenges. Template banks require extensive computational resources to cover the multi-dimensional parameter space of source properties, and the approach becomes less efficient when dealing with poorly modeled or unexpected signal morphologies. Additionally, non-stationary instrumental noise and transient artifacts can trigger false alarms, necessitating sophisticated vetting procedures \cite{cain2017methods}.

The emergence of machine learning techniques in gravitational wave astronomy has shown promising potential for addressing these limitations \cite{george2018deep, gabbard2018matching}. Deep learning approaches offer several advantages: they can learn complex patterns directly from data without requiring explicit waveform models, adapt to varying noise conditions, and potentially identify novel signal classes. Previous studies have explored convolutional neural networks for binary classification \cite{george2018deep}, recurrent networks for time-series analysis \cite{schafer2020training}, and generative models for parameter estimation \cite{green2020gravitational}. However, most existing approaches treat gravitational wave detection as a straightforward time-series classification problem, potentially overlooking the rich time-frequency structure inherent in chirping signals.

Gravitational wave signals from compact binary coalescences exhibit characteristic time-frequency evolution, with frequency increasing as the binary components spiral inward toward merger. This "chirp" behavior is optimally captured through time-frequency analysis techniques. The Continuous Wavelet Transform (CWT) provides an ideal framework for decomposing gravitational wave strain data, offering superior time-frequency resolution compared to short-time Fourier transforms and maintaining the temporal localization essential for transient signal detection \cite{chatterji2004multiresolution}. Recent studies have demonstrated the effectiveness of CWT preprocessing for gravitational wave analysis \cite{powell2015classification}. While unsupervised learning approaches including autoencoders have been applied to gravitational wave detection \cite{torres2024anomaly, lopez2022deep}, the specific integration of CWT preprocessing with LSTM autoencoder architectures for gravitational wave detection has not been systematically investigated.

Autoencoder networks, a class of unsupervised learning models, excel at learning compressed representations of complex data while preserving essential features \cite{hinton2006reducing}. Long Short-Term Memory (LSTM) autoencoders extend this capability to sequential data, capturing long-range temporal dependencies crucial for modeling the extended duration of gravitational wave signals \cite{hochreiter1997long}. The reconstruction-based nature of autoencoders provides an intuitive framework for anomaly detection: signals that deviate significantly from learned noise patterns can be identified as potential gravitational wave candidates.

This work investigates the combination of CWT preprocessing with LSTM autoencoder architecture for gravitational wave detection. The method leverages the optimal time-frequency representation provided by wavelets while utilizing the temporal modeling capabilities of recurrent neural networks. We demonstrate the effectiveness of this approach on realistic synthetic datasets incorporating binary black hole merger signals embedded in colored Gaussian noise representative of Advanced LIGO sensitivity curves. Our results achieve 90.6\% precision, exceeding the stringent detection requirements necessary for confident gravitational wave identification.

The paper is organized as follows: Section 2 describes the methodology, including CWT preprocessing and LSTM autoencoder architecture. Section 3 presents the synthetic data generation procedure and training protocol. Section 4 analyzes the results and compares performance with baseline approaches. Section 5 discusses implications for real-world applications and future directions toward implementation with actual LIGO data.

\section{Methodology}

\subsection{Continuous Wavelet Transform Preprocessing}

The Continuous Wavelet Transform provides optimal time-frequency decomposition for gravitational wave signals, preserving both temporal localization and frequency resolution essential for chirp detection. For a given strain time series $h(t)$, the CWT is defined as:

\begin{equation}
W(a,b) = \frac{1}{\sqrt{a}} \int_{-\infty}^{\infty} h(t) \psi^*\left(\frac{t-b}{a}\right) dt
\end{equation}

where $\psi(t)$ is the mother wavelet, $a$ is the scale parameter inversely related to frequency, $b$ is the translation parameter corresponding to time, and $*$ denotes complex conjugation.

We employ the Morlet wavelet as the mother wavelet due to its optimal time-frequency localization properties and resemblance to gravitational wave chirp morphology:

\begin{equation}
\psi(t) = \pi^{-1/4} e^{i\omega_0 t} e^{-t^2/2}
\end{equation}

where $\omega_0 = 6$ provides the optimal trade-off between time and frequency resolution for gravitational wave analysis. The CWT is computed over logarithmically spaced scales corresponding to frequencies from 20 Hz to 512 Hz, matching the sensitive frequency band of Advanced LIGO detectors.

The resulting time-frequency representation forms a 2D scalogram $|W(a,b)|^2$ that captures the characteristic frequency evolution of gravitational wave chirps. This scalogram serves as input to the subsequent neural network architecture, providing rich feature representations that preserve both the temporal evolution and spectral content of potential signals.

\subsection{LSTM Autoencoder Architecture}

The core detection system employs a Long Short-Term Memory autoencoder designed to learn compressed representations of gravitational wave time-frequency patterns. The architecture consists of three main components: an encoder network, a bottleneck layer, and a decoder network.

\subsubsection{Encoder Network}

The encoder processes the CWT scalogram through a series of LSTM layers with progressively reducing hidden dimensions:

\begin{equation}
h_t^{(l)} = \text{LSTM}^{(l)}(x_t^{(l)}, h_{t-1}^{(l)})
\end{equation}

where $h_t^{(l)}$ represents the hidden state at time $t$ for layer $l$, and $x_t^{(l)}$ is the input at time $t$ for layer $l$. The encoder employs three LSTM layers with hidden dimensions [128, 64, 32], capturing hierarchical temporal patterns at multiple scales.

Each LSTM cell implements the standard gating mechanism:

\begin{align}
f_t &= \sigma(W_f \cdot [h_{t-1}, x_t] + b_f) \\
i_t &= \sigma(W_i \cdot [h_{t-1}, x_t] + b_i) \\
\tilde{C}_t &= \tanh(W_C \cdot [h_{t-1}, x_t] + b_C) \\
C_t &= f_t \cdot C_{t-1} + i_t \cdot \tilde{C}_t \\
o_t &= \sigma(W_o \cdot [h_{t-1}, x_t] + b_o) \\
h_t &= o_t \cdot \tanh(C_t)
\end{align}

where $\sigma$ denotes the sigmoid activation function, $W$ and $b$ represent learned weight matrices and bias vectors, and $f_t$, $i_t$, $o_t$ are the forget, input, and output gates respectively.

\subsubsection{Bottleneck Representation}

The encoder output feeds into a dense bottleneck layer with 16 neurons, forcing the network to learn a compact latent representation of the input signal. This compression step ensures that only the most salient features necessary for reconstruction are preserved, naturally filtering out noise components that cannot be efficiently encoded.

\subsubsection{Decoder Network}

The decoder mirrors the encoder architecture with LSTM layers of dimensions [32, 64, 128], reconstructing the original CWT scalogram from the compressed representation. The final layer employs linear activation to match the continuous-valued input domain.

\subsection{Training Protocol}

\subsubsection{Loss Function}

The network is trained using mean squared error (MSE) loss between the input and reconstructed scalograms:

\begin{equation}
\mathcal{L} = \frac{1}{N} \sum_{i=1}^{N} ||X_i - \hat{X}_i||^2
\end{equation}

where $X_i$ represents the input scalogram, $\hat{X}_i$ is the reconstruction, and $N$ is the batch size.

\subsubsection{Optimization}

Training employs the Adam optimizer with an initial learning rate of $10^{-3}$, exponential decay rate of 0.95 every 10 epochs, and gradient clipping at magnitude 1.0 to ensure stable convergence. The network trains for 100 epochs with early stopping based on validation loss plateau.

\subsection{Detection Strategy}

Gravitational wave detection operates on the principle that signals containing true astrophysical events will exhibit higher reconstruction error compared to pure noise segments. For each test sample, we compute the reconstruction error:

\begin{equation}
E = ||X - \hat{X}||^2
\end{equation}

Samples with reconstruction error exceeding a predetermined threshold $\tau$ are classified as potential gravitational wave candidates. The threshold is optimized using precision-recall analysis on validation data to achieve the desired balance between detection sensitivity and false alarm rate.

\subsection{Performance Metrics}

Model performance is evaluated using standard binary classification metrics:

\begin{align}
\text{Precision} &= \frac{TP}{TP + FP} \\
\text{Recall} &= \frac{TP}{TP + FN} \\
\text{F1-Score} &= \frac{2 \cdot \text{Precision} \cdot \text{Recall}}{\text{Precision} + \text{Recall}}
\end{align}

where $TP$, $FP$, and $FN$ represent true positives, false positives, and false negatives respectively. Additionally, we compute the Area Under the Precision-Recall Curve (AUPRC) as a threshold-independent performance measure particularly relevant for imbalanced datasets typical in gravitational wave detection scenarios.

\section{Data Generation and Experimental Setup}

\subsection{Synthetic Gravitational Wave Signal Generation}

We generate realistic synthetic gravitational wave signals representing binary black hole (BBH) coalescences using post-Newtonian waveform approximations. The gravitational wave strain is modeled as:

\begin{equation}
h(t) = h_+(t) \cos(2\psi) + h_\times(t) \sin(2\psi)
\end{equation}

where $h_+(t)$ and $h_\times(t)$ are the plus and cross polarizations, and $\psi$ is the polarization angle randomly sampled from $[0, 2\pi]$.

\subsubsection{Chirp Signal Modeling}

The frequency evolution follows the post-Newtonian expansion for the inspiral phase:

\begin{equation}
f(t) = f_0 \left(\frac{\tau}{\tau_0}\right)^{-3/8}
\end{equation}

where $f_0 = 35$ Hz is the initial frequency, $\tau = t_c - t$ is the time to coalescence, and $\tau_0$ is the initial time to coalescence. The instantaneous phase evolves as:

\begin{equation}
\phi(t) = 2\pi \int_0^t f(t') dt'
\end{equation}

The amplitude incorporates realistic scaling with chirp mass $\mathcal{M}_c$ and luminosity distance $D_L$:

\begin{equation}
A(t) = \frac{4G^{5/3}}{c^3} \frac{(2\pi f(t))^{2/3} \mathcal{M}_c^{5/3}}{D_L}
\end{equation}

where $G$ is the gravitational constant and $c$ is the speed of light.

\subsubsection{Parameter Distributions}

Binary system parameters are drawn from astrophysically motivated distributions:

\begin{itemize}
\item \textbf{Component masses}: $m_1, m_2 \sim \mathcal{U}(10, 80) M_\odot$ with $m_1 \geq m_2$
\item \textbf{Distance}: $D_L \sim \mathcal{U}(100, 1000)$ Mpc
\item \textbf{Inclination}: $\cos(\iota) \sim \mathcal{U}(-1, 1)$ (isotropic distribution)
\item \textbf{Sky location}: Uniform distribution over the celestial sphere
\item \textbf{Coalescence time}: Randomly placed within the 4-second observation window
\end{itemize}

The chirp mass and symmetric mass ratio are derived as:

\begin{align}
\mathcal{M}_c &= \frac{(m_1 m_2)^{3/5}}{(m_1 + m_2)^{1/5}} \\
\eta &= \frac{m_1 m_2}{(m_1 + m_2)^2}
\end{align}

\subsection{Noise Modeling}

\subsubsection{Advanced LIGO Noise Characteristics}

We model realistic detector noise using the Advanced LIGO design sensitivity curve, incorporating both fundamental noise sources and instrumental artifacts. The power spectral density (PSD) follows the analytical approximation:

\begin{equation}
S_n(f) = S_0 \left[ \left(\frac{f}{f_0}\right)^{-4.14} + 5 + 3\left(\frac{f}{f_0}\right)^2 \right]
\end{equation}

where $S_0 = 10^{-49}$ Hz$^{-1}$ and $f_0 = 215$ Hz represent the characteristic noise amplitude and knee frequency respectively.

\subsubsection{Colored Gaussian Noise Generation}

Colored Gaussian noise matching the LIGO PSD is generated using the frequency-domain method:

\begin{enumerate}
\item Generate white Gaussian noise $\tilde{n}_{\text{white}}(f)$ in the frequency domain
\item Apply spectral shaping: $\tilde{n}(f) = \tilde{n}_{\text{white}}(f) \sqrt{S_n(f)}$
\item Transform to time domain: $n(t) = \text{IFFT}[\tilde{n}(f)]$
\end{enumerate}

This procedure ensures that the generated noise accurately reproduces the frequency-dependent sensitivity characteristics of Advanced LIGO detectors.

\subsection{Dataset Construction}

\subsubsection{Signal-to-Noise Ratio Distribution}

Signals are injected into noise with optimal signal-to-noise ratios (SNRs) distributed according to:

\begin{equation}
\rho_{\text{opt}} = \sqrt{4 \int_{f_{\text{low}}}^{f_{\text{high}}} \frac{|\tilde{h}(f)|^2}{S_n(f)} df}
\end{equation}

where $\tilde{h}(f)$ is the Fourier transform of the strain signal, and the integration limits span the detector's sensitive frequency band [20, 512] Hz. SNRs are drawn from the range [8, 25], representing the spectrum from threshold-level to highly significant detections.

\subsubsection{Dataset Composition}

The complete dataset comprises:

\begin{itemize}
\item \textbf{Training set}: 80,000 samples (40,000 pure noise + 40,000 signal+noise)
\item \textbf{Validation set}: 10,000 samples (5,000 pure noise + 5,000 signal+noise)
\item \textbf{Test set}: 10,000 samples (5,000 pure noise + 5,000 signal+noise)
\end{itemize}

Each sample represents a 4-second time series sampled at 512 Hz, yielding 2048 data points per observation. This duration captures the entire inspiral phase for the considered mass range while maintaining computational tractability.

\subsection{Experimental Configuration}

\subsubsection{Computational Infrastructure}

Training and evaluation are performed on a system equipped with:

\begin{itemize}
\item \textbf{CPU}: Intel i7-12700K (8 cores, 16 threads)
\item \textbf{GPU}: NVIDIA RTX 3080 (10 GB VRAM)
\item \textbf{Memory}: 32 GB DDR4-3200
\item \textbf{Framework}: PyTorch 2.0 with CUDA acceleration
\end{itemize}

\subsubsection{Training Hyperparameters}

The optimal hyperparameter configuration, determined through systematic grid search, includes:

\begin{itemize}
\item \textbf{Batch size}: 64 samples
\item \textbf{Learning rate}: $10^{-3}$ with exponential decay (factor 0.95, every 10 epochs)
\item \textbf{Dropout rate}: 0.2 (applied to LSTM layers during training)
\item \textbf{Weight decay}: $10^{-5}$ for L2 regularization
\item \textbf{Gradient clipping}: Maximum norm of 1.0
\item \textbf{Early stopping}: Patience of 15 epochs based on validation loss
\end{itemize}

\subsubsection{Cross-Validation Strategy}

To ensure robust performance estimates, we employ 5-fold cross-validation on the combined training and validation sets. Each fold maintains the 50-50 balance between signal and noise samples, and final performance metrics represent the mean and standard deviation across all folds.

\subsection{Baseline Comparisons}

We compare the CWT-LSTM autoencoder against several baseline approaches:

\begin{itemize}
\item \textbf{Matched filtering}: Template-based detection using PyCBC implementation
\item \textbf{Convolutional Neural Network}: 1D CNN with temporal convolutions
\item \textbf{Standard LSTM classifier}: Direct time-series classification without CWT preprocessing
\item \textbf{Spectral analysis}: Power spectral density-based anomaly detection
\end{itemize}

All baseline methods are trained and evaluated on identical datasets to ensure fair comparison, with hyperparameters optimized independently for each approach.

% Include the auto-updating results section
\input{sections/results}

\section{Discussion}

\subsection{Significance of Results}

The CWT-LSTM autoencoder achieves 90.6\% precision in gravitational wave detection, representing a substantial advancement toward the stringent requirements of operational observatories. This performance level exceeds the false alarm rates necessary for confident gravitational wave identification, where precision above 90\% typically corresponds to false alarm rates acceptable for scientific publication and follow-up observations.

The achieved results are particularly significant when considered in the context of LIGO's detection thresholds. Current LIGO pipelines require signal-to-noise ratios corresponding to false alarm rates less than 1 per year for confident detections. Our precision of 90.6\% suggests that the CWT-LSTM approach could potentially meet these requirements when deployed on real detector data, though validation on actual LIGO strain measurements remains essential.

\subsection{Methodological Contributions}

\subsubsection{Time-Frequency Preprocessing Innovation}

The integration of Continuous Wavelet Transform preprocessing with LSTM autoencoders represents an approach that leverages the complementary strengths of both paradigms. Traditional matched filtering excels at detecting signals with precisely known waveforms but struggles with computational scalability and unexpected signal morphologies. Deep learning approaches offer adaptability and pattern recognition capabilities but typically ignore the rich time-frequency structure inherent in gravitational wave signals.

The CWT preprocessing preserves the essential chirp characteristics while providing a standardized input representation that facilitates neural network training. Our ablation studies demonstrate that removing CWT preprocessing reduces precision by 16.2 percentage points, confirming the critical importance of this methodological choice.

\subsubsection{Autoencoder Architecture Advantages}

The autoencoder framework provides several advantages over conventional classification approaches. By learning to reconstruct normal noise patterns, the model naturally identifies anomalous signals without requiring explicit labeled examples of all possible gravitational wave types. This unsupervised learning component, building upon recent work in autoencoder-based gravitational wave detection \cite{torres2024anomaly, lopez2022deep}, enhances the method's potential for discovering unexpected signal classes or novel astrophysical phenomena. Our specific contribution lies in the systematic integration of time-frequency preprocessing through CWT with the temporal modeling capabilities of LSTM autoencoders.

The LSTM architecture captures the temporal evolution essential for gravitational wave detection while maintaining computational efficiency. The hierarchical encoding strategy enables the model to learn features at multiple temporal scales, from short-duration transients to extended inspiral signals lasting several seconds.

\subsection{Comparison with Existing Methods}

\subsubsection{Matched Filtering Performance}

While matched filtering achieves comparable F1-scores (0.788 vs 0.778), the CWT-LSTM autoencoder's superior precision (90.6\% vs 82.1\%) represents a significant practical advantage. In gravitational wave astronomy, false alarms impose substantial costs in terms of follow-up observations, computational resources, and scientific credibility. The 8.5 percentage point improvement in precision could translate to meaningful reductions in false alarm rates for operational detection systems.

Moreover, matched filtering requires extensive template banks covering the multidimensional parameter space of possible sources. The CWT-LSTM approach learns signal characteristics directly from data, potentially offering computational advantages and enhanced adaptability to unexpected signal types.

\subsubsection{Alternative Deep Learning Approaches}

The comparison with standard neural network architectures (1D CNN, LSTM classifier) demonstrates the importance of the combined CWT-autoencoder methodology. The precision improvements over alternative approaches validate the architectural choices made in this work.

These performance differences highlight the value of domain-specific preprocessing and appropriate neural network design for gravitational wave detection applications.

\subsection{Computational Efficiency}

The 2.3-millisecond inference time per 4-second segment enables real-time analysis of gravitational wave data streams, a critical requirement for operational detection systems. The training time of 6 hours represents a reasonable computational investment that could be easily accommodated in research or operational environments.

The model's memory requirements (8.2 GB peak) remain within the capacity of standard GPU hardware, suggesting feasible deployment without specialized computational infrastructure. This accessibility could facilitate widespread adoption and further development by the gravitational wave community.

\subsection{Limitations and Future Directions}

\subsubsection{Synthetic Data Constraints}

The primary limitation of this work lies in its reliance on synthetic data generated from theoretical waveform models. Real gravitational wave detection faces additional challenges including:

\begin{itemize}
\item Non-stationary detector noise with complex spectral characteristics
\item Instrumental glitches and environmental artifacts
\item Unknown signal morphologies from exotic sources
\item Calibration uncertainties and systematic errors
\end{itemize}

Validation on actual LIGO data represents the critical next step for establishing practical applicability. The transition from synthetic to real data often reveals unexpected challenges that cannot be fully anticipated through simulation alone.

\subsubsection{Signal Diversity}

Our synthetic dataset focuses primarily on binary black hole coalescences with specific parameter ranges. Gravitational wave astronomy encompasses a broader spectrum of sources including binary neutron stars, neutron star-black hole mergers, and potentially exotic phenomena like cosmic strings or primordial black holes. Expanding the training dataset to include greater signal diversity would enhance the model's generalization capabilities.

\subsubsection{Detector Network Analysis}

Current gravitational wave detection relies on coincident observations across multiple detectors to reject local noise artifacts and improve source localization. This work focuses on single-detector analysis, while operational systems require coherent analysis across the LIGO-Virgo-KAGRA network. Extending the CWT-LSTM approach to multi-detector coincidence analysis represents an important avenue for future development.

\subsection{Path Toward Real-World Implementation}

\subsubsection{LIGO Data Integration}

The next critical milestone involves applying the CWT-LSTM autoencoder to actual LIGO strain data available through the Gravitational Wave Open Science Center. This validation step will reveal the method's performance on real noise characteristics and known gravitational wave events from the GWTC catalogs.

Initial implementation should focus on reprocessing archived data from completed observing runs, allowing direct comparison with established detection pipelines and validation against confirmed gravitational wave events.

\subsubsection{Online Detection Pipeline}

For operational deployment, the CWT-LSTM autoencoder would need integration into real-time detection pipelines processing continuous data streams. This integration requires:

\begin{itemize}
\item Streaming data preprocessing and buffering systems
\item Real-time CWT computation with optimized algorithms
\item Distributed computing infrastructure for parallel processing
\item Alert generation and follow-up coordination systems
\end{itemize}

\subsubsection{Collaborative Development}

The gravitational wave community's collaborative nature suggests that open-source implementation and community validation would accelerate adoption. Making the trained models, preprocessing tools, and evaluation frameworks publicly available could foster further research and refinement.

\subsection{Broader Implications}

\subsubsection{Multi-Messenger Astronomy}

Improved gravitational wave detection capabilities directly impact multi-messenger astronomy by enabling more reliable electromagnetic follow-up observations. Higher precision detection systems reduce false alarms that consume limited telescope resources and improve the scientific return from coordinated observational campaigns.

\subsubsection{Fundamental Physics}

Enhanced detection sensitivity could facilitate discovery of weaker gravitational wave sources and enable more precise tests of general relativity. The CWT-LSTM approach's potential for identifying unexpected signal types could contribute to searches for exotic phenomena beyond the standard model.

\subsubsection{Discovery Potential for Unknown Signal Types}

A fundamental advantage of the CWT-LSTM autoencoder approach lies in its unsupervised learning paradigm, which enables detection of gravitational wave signals without prior knowledge of their morphology. This capability represents a paradigm shift from traditional matched filtering techniques that require pre-computed template banks based on theoretical waveform models.

\textbf{Template-Free Detection:} Unlike matched filtering, which can only detect signals that closely match existing templates, the autoencoder learns to identify any deviation from the learned noise distribution. This template-free approach eliminates the computational burden of generating and searching through extensive waveform libraries while simultaneously opening the possibility of detecting entirely unexpected signal types.

\textbf{Discovery of Exotic Sources:} The method's sensitivity to anomalous patterns positions it as an ideal tool for discovering gravitational waves from exotic astrophysical sources not covered by current template banks. Potential discoveries could include:

\begin{itemize}
\item \textbf{Cosmic strings}: Theoretical one-dimensional defects in spacetime that could produce distinctive gravitational wave signatures
\item \textbf{Primordial black holes}: Remnants from the early universe with potentially unique merger characteristics
\item \textbf{Exotic compact objects}: Hypothetical alternatives to black holes, such as boson stars or gravastars
\item \textbf{Modified gravity signatures}: Deviations from General Relativity predictions that manifest in unexpected waveform features
\item \textbf{Unknown physics}: Completely novel phenomena that current theoretical models cannot predict
\end{itemize}

\textbf{Complementary Discovery Strategy:} Rather than replacing matched filtering, the CWT-LSTM autoencoder provides a complementary "blind search" capability that could identify candidate events for subsequent detailed analysis. This two-stage approach maximizes both sensitivity to known signals and discovery potential for unknown phenomena.

\textbf{Scientific Impact:} The ability to detect unexpected gravitational wave signatures could revolutionize our understanding of fundamental physics and astrophysics. Historical precedent in astronomy demonstrates that the most significant discoveries often come from detecting the unexpected—from pulsars to gamma-ray bursts to dark energy. The CWT-LSTM autoencoder provides gravitational wave astronomy with a similar capability for serendipitous discovery.

This discovery potential represents one of the most compelling advantages of unsupervised approaches in gravitational wave detection, offering the possibility of uncovering new physics that would remain hidden to template-based searches.

\subsubsection{Machine Learning in Physics}

This work demonstrates the value of domain-specific preprocessing and architecture design in scientific machine learning applications. The successful integration of established signal processing techniques (CWT) with modern deep learning illustrates a productive path for physics applications where domain knowledge can guide neural network design.

\subsection{Conclusion}

The CWT-LSTM autoencoder represents a significant advancement in gravitational wave detection methodology, achieving 90.6\% precision on realistic synthetic data. The combination of time-frequency preprocessing with recurrent neural networks provides a powerful framework that leverages both domain expertise and modern machine learning capabilities.

While validation on real LIGO data remains essential, the demonstrated performance suggests substantial potential for operational gravitational wave detection systems. The method's computational efficiency, superior precision compared to existing approaches, and potential for discovering unexpected signal types position it as a valuable addition to the gravitational wave detection toolkit.

Future work will focus on real data validation, multi-detector analysis, and integration with operational detection pipelines to fully realize the potential of this approach for advancing gravitational wave astronomy.

\section{Acknowledgments}

The author thanks the gravitational wave community for making LIGO data and analysis tools publicly available through the Gravitational Wave Open Science Center. Special appreciation goes to the developers of PyTorch, PyWavelets, and other open-source libraries that enabled this research. The author acknowledges Portland Community College for providing a supportive environment for physics education and research pursuits, and Adobe for fostering innovation and scientific exploration.

This research made use of computational resources provided by NVIDIA through their academic GPU programs. The author also acknowledges the LIGO Scientific Collaboration and Virgo Collaboration for their groundbreaking work in gravitational wave detection that motivated and informed this study.

\bibliography{references}
\bibliographystyle{iopart-num}

\end{document}
